---
title: "BranchRename: Updating the default branch in Bioconductor"
author: "Marcel Ramos"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{BranchRename FAQs}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: true
---

# GitHub Installation

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("Bioconductor/BranchRename")
```

## Loading the package

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(BranchRename)
```

# A Brief Description

`BranchRename` offers both organization and user-level tools for renaming
default branches on GitHub. It also interacts with local repositories that
need to be sync'ed with remote GitHub repositories.

# Setup

## Authentication

First, on GitHub, create a fine-grained Personal Access Token (PAT) under User
\> Settings \> Developer Settings \> PATs \> Fine-grained tokens.

When generating a new token, select `Bioconductor` (or organization) as the
resource owner, then 'All repositories' and under 'Repository permissions'
select 'Contents' \> 'Read and Write' and also select 'Read and Write' under
'Administration'. It is recommended to use `gitcreds::gitcreds_set()` to store
'the fine-grained PAT and to save the token using Seahorse (Ubuntu) or other
credentials management application native to your operating system.

**Note**. Depending on the organization's settings, R / Bioconductor packages
hosted on GitHub may also be updated with a classic PAT. These commands only
work if the user has admin access to the organization's GitHub account.

# Steps

## Identify repositories

The Bioconductor organization on GitHub hosts repositories that may also be
packages in Bioconductor. To identify packages containing the old default branch
name use the `packages_with_default_branch()` function. To identify repositories
that use the old default branch use the `repos_with_default_branch()` function.

**Note**. The functions by default identify repositories with both `master` and
`main` as the default branch.

## Rename branch in GitHub repository

### Single package

The `rename_branch_to_devel` function allows users to rename the default branch
of a single package hosted by an organization on GitHub. For example, here we 
attempt to change the default branch from `master` to `devel` and ensure that
upstream tracking is set to `origin/devel` or `upstream/devel`:

```r
rename_branch_to_devel(
    package_name = "SummarizedExperiment",
    from_branch = "master",
    org = "Bioconductor",
    set_upstream = "origin/devel",
    clone = TRUE
)
```

Note that `origin` and `upstream` refer to the GitHub and Bioconductor git
(i.e., on [git.bioconductor.org][1]) repositories, respectively.

### Batch of packages

To rename branches for multiple packages, it is first recommended to use the
`packages_with_default_branch()` to identify the groups of packages under
the organization that need to have their branches renamed. The output of that
function will provide a named character vector, e.g., `c(SummarizedExperiment =
"master", GenomicRanges = "master")` where the values are the default branch
names and the names are the package names.

```r
pkgs <- packages_with_default_branch()
```

With the named character vector, one can use the `rename_branch_packages`
function to perform the rename:

```r
rename_branch_packages(packages = pkgs)
```

Similar operations can be performed with both `repos_with_default_branch` and
`rename_branch_repos`.

Both sets of functions work similarly but differ by package status. Bioconductor
packages are expected to have an `upstream` git remote that points to
[git.bioconductor.org][1].

## Local repository update

For local repositores that were missed, e.g., when the `rename_branch_to_devel`
or the `rename_branch_packages` functions were run with `clone = TRUE` in a
separate directory, the old repositories would need the local branches to be
updated to the new name. The `update_local_repo` function allows the user
to perform the rename operation locally and sync the repository to the
GitHub remote repository (in other words set the `HEAD` to track
`origin/devel`).

```r
update_local_repo(
    repo_dir = "~/bioc/SummarizedExperiment",
    new_branch = "devel",
    set_upstream = "origin/devel"
)
```

The function `update_local_repos` allows these type of updates for multiple
repositories within a given directory (`basedir`).

### Utilties

The `add_bioc_remote` is a convenience function that creates an `upstream`
remote for a given local git repository. This is useful only for packages
that are in [git.bioconductor.org][1] whose local repositories do not have
an `upstream` remote. The `add_bioc_remote()` function is equivalent to:

```sh
git remote add upstream git@git.bioconductor.org:packages/<MyPackage>
```

[1]: https://git.bioconductor.org/

# Session Info

```{r}
sessionInfo()
```